The fraction of polar aligned circumbinary disks

### 2025-02-07 src/tex/ms.tex
$h$ is constant language change
Fix jmax math

\subsection{Self-gravitation}
\label{subsec:toomre}
Just as the disk-breaking framework in Section \ref{subsec:breaks} places an upper limit on the outer radius of an extended disk, we can use the gravitational instability criterion to place a lower limit on that outer radius for constant $j$. In this section we will calculate the maximum $j$ that is allowed for a disk with a given radial extent before it becomes gravitationally unstable. This discussion applies to any disk that one may wish to describe with the dimensionless relative angular momentum $j$, not just the misaligned disks we have focused on so far. The criterion for stability against self-gravitation \citep{toomre1964} can be written
\begin{equation}
    \label{eq:toomre}
    \frac{hM_\text{b}}{\pi r^2 \Sigma} > 1\, ,
\end{equation}
where $h\equiv H/r$ is the aspect ratio of the disk and $\Sigma$ is the surface density. We will assume $h$ is constant. If $\Sigma = \Sigma_\text{in} (r/r_\text{in})^{-p}$ and $p>0$, then the angular momentum of the ring (in the limit $M_\text{r} \ll M_\text{b}$) is
\begin{align}
    J_\text{r} &= \int_{r_\text{in}}^{r_\text{out}} 2\pi r \Sigma  \sqrt{GM_\text{b}r}\,dr \, , \nonumber \\
    & = \frac{2\pi\sqrt{GM_\text{b}}}{5/2 - p} \Sigma_\text{in} r_\text{in}^{5/2} \left( \left( \frac{r_\text{out}}{r_\text{in}} \right)^{5/2 - p}- 1\right)\, \label{eq:Jr}.
\end{align}

@@

The value of $p$ is typically taken to be less than 2, so the instability onset occurs at the inner edge of the disk. We use equation (\ref{eq:toomre2}) to find a maximum value for $\Sigma_\text{in}$:
\begin{equation}
    \Sigma_\text{max} = \frac{hM_\text{b}}{\pi r_\text{in}^2}\, .
\end{equation}

The maximum value of $j$ to be stable against self-gravity is
\begin{equation}
    \label{eq:jmax-shallow}
    j_\text{max} = \frac{2h}{f (1-f) (5/2 - p)\sqrt{1-e_\text{b}^2}} \sqrt{\frac{r_\text{in}}{a_\text{b}}} \left( \left( \frac{r_\text{out}}{r_\text{in}} \right)^{5/2 - p}- 1\right)\, .
\end{equation}

If we consider a $p=1$ disk orbiting an equal mass binary with $e_\text{b}=0.4$, $h=0.01$, and $r_\text{in}/a_\text{b}=2.5$, then for a relatively narrow ring ($r_\text{out}/r_\text{in} = 2$), we calculate $j_\text{max} = 0.17$. If the disk is allowed to be more extended (in this case $r_\text{out}/r_\text{in} = 10$), then we find $j_\text{max} = 2.8$.

>>>>

\subsection{Self-gravitation}
\label{subsec:toomre}
Just as the disk-breaking framework in Section \ref{subsec:breaks} places an upper limit on the outer radius of an extended disk, we can use the gravitational instability criterion to place a lower limit on that outer radius for constant $j$. In this section we will calculate the maximum $j$ that is allowed for a disk with a given radial extent before it becomes gravitationally unstable. This discussion applies to any disk that one may wish to describe with the dimensionless relative angular momentum $j$, not just the misaligned disks we have focused on so far. The criterion for stability against self-gravitation \citep{toomre1964} can be written
\begin{equation}
    \label{eq:toomre}
    \frac{hM_\text{b}}{\pi r^2 \Sigma} > 1\, ,
\end{equation}
where $h\equiv H/r$ is the aspect ratio of the disk that is assumed to be constant and $\Sigma$ is the surface density. If $\Sigma = \Sigma_\text{in} (r/r_\text{in})^{-p}$ and $p>0$, then the angular momentum of the ring (in the limit $M_\text{r} \ll M_\text{b}$) is
\begin{align}
    J_\text{r} &= \int_{r_\text{in}}^{r_\text{out}} 2\pi r \Sigma  \sqrt{GM_\text{b}r}\,dr \, , \nonumber \\
    & = \frac{2\pi\sqrt{GM_\text{b}}}{5/2 - p} \Sigma_\text{in} r_\text{in}^{5/2} \left( \left( \frac{r_\text{out}}{r_\text{in}} \right)^{5/2 - p}- 1\right)\, \label{eq:Jr}.
\end{align}

@@

The value of $p$ is typically taken to be less than 2, so the instability onset occurs at the outer edge of the disk. We use equation (\ref{eq:toomre2}) to find a maximum value for $\Sigma_\text{in}$:
\begin{equation}
    \Sigma_\text{max} = \frac{hM_\text{b}}{\pi r_\text{in}^2 (r_\text{out}/r_\text{in})^{2-p}}\, .
\end{equation}

The maximum value of $j$ to be stable against self-gravity is
\begin{multline}
    \label{eq:jmax-shallow}
    j_\text{max} = \frac{2h}{f (1-f) (5/2 - p)\sqrt{1-e_\text{b}^2}}\\ \sqrt{\frac{r_\text{in}}{a_\text{b}}} \left(\frac{r\text{out}}{r_\text{in}}\right)^{p-2} \left( \left( \frac{r_\text{out}}{r_\text{in}} \right)^{5/2 - p}- 1\right)\, .
\end{multline}

If we consider a $p=1$ disk orbiting an equal mass binary with $e_\text{b}=0.4$, $h=0.01$, and $r_\text{in}/a_\text{b}=2.5$, then for a relatively narrow ring ($r_\text{out}/r_\text{in} = 2$), we calculate $j_\text{max} = 0.08$. If the disk is allowed to be more extended (in this case $r_\text{out}/r_\text{in} = 10$), then we find $j_\text{max} = 0.28$.
